---
title: "Kronos"
author: "Sarah-Jane Leigh & Thomaz F.S. Bastiaanssen"
date: "2022-10-26"
output:
  md_document:
    variant: gfm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Please note this package is under active construction.

# Kronos

If you use this software, please cite our work:

*#include citation here with DOI*

The following document is adapted from the supplementary materials for this manuscript.

## 0. Introduction

Here, we will demonstrate how to use the Kronos package to assess circadian rhythms in biological data collected over the day. For this demonstration we have adapted some data from our laboratory, currently undergoing peer review (reference to come).

We will demonstrate three common examples of experimental design currently encountered in circadian rhythms analysis:

1.  Circadian rhythm analysis of 1 variable over a 24-hour period

2.  Circadian rhythm analysis of 1 variable over a 24-hour period with two or more treatment groups

3.  Circadian rhythm analysis of omics data over a 24-hour period with two or more treatment groups

All R code used to transform, reorganise and plot the data is also shown below to hopefully provide a toolkit for aspiring and veteran bioinformaticians alike. It should be noted that the analysis performed here may not perfectly correspond to that performed in the published manuscript: some variables in the datasets provided have been manipulated to better demonstrate the functionality of this package.

*At the end of this document, we have included a few excursions into more advanced subjects we find useful, but that did not necessarily fit in with the mainline analysis.*

### Code chunk: Install Kronos and load our libraries

```{r load, message=FALSE}

#Install kronos
library(devtools)
#install_github("thomazbastiaanssen/kronos") 

#Load relevant packages
library(kronos)
library(tidyverse)
library(ggplot2)

#Load prepared data stored in Kronos library
data("kronos_demo")

```

\newpage

### Code chunk: Load our data tables

Data should be prepared in "long form" for use in Kronos; that is with values repeating in the "Timepoint" column, which defines when data was collected during the period, here a 24-hour cycle.

Our package includes example datasets that we will use in this tutorial that are pre-formatted. You can rearrange your data into long form using pivot_longer() or gather() in tidyr.

Since we're using prepared data, we already loaded it using `data(kronos_demo)`. You can see how an example of how the data is prepared below:

```{r showdata, echo=FALSE}

head(groupdata)

```

Here we have prepared the omics dataset with a separate metadata file as is common when working with omics datasets. A metadata file can be generated using select() in tidyr, or metadata and data can be combined using inner_join if they contain an identical column (both name and contents). The omics dataset used here has been central-log transformed to account for its compositional nature (see our guide <https://arxiv.org/abs/2207.12475> for easy centred-log transformation).

## 1. Statistical Approach

*To be completed*

## 2. Analysing Rhythmicity in a Single Group

We will start with the most simple example: analysing circadian rhythmicity in a single experimental group for one outcome variable of interest. For this we use the kronos function:

```{r kronos1group, message=FALSE, warning=FALSE, cache=FALSE}

output <- kronos(formula = Variable_1 ~ time(Timepoint), 
                 data = onevariable, period = 24, 
                 verbose = T, pairwise = F)

```

Here we use the formula `Outcome Variable ~ time(Time Variable)`, which is the most simple model used by the Kronos function. We specify the period as 24 (this can be adjusted as appropriate for the data analysed). By selecting `verbose=T`, you will be able to see the models run by the kronos function: this becomes increasingly useful when you run more complex models. Finally we select `pairwise=F` here, as there are no groups to compare for differences in rhythms.

The kronos function creates a kronosOut object, containing several pieces of data that can be accessed using handy getter functions, which we will describe below:

1). The *getKronos_input()* function fetches the data that the model is based on, as well as the calculated cosine and sine components.

```{r showinput, echo=FALSE}

head(getKronos_input(output))
```

2). The *getKronos_fit()* function fetches the key details for the generated model that may be useful for prediction, modelling and other statistical applications.

```{r showfit}

getKronos_fit(output)

```

3). The *getKronos_trace()* functioncontains all the data required for graphing the sinusoid curve, which can either be used in our custom ggplot2 functions, or can be used in other graphing packages. The *y_hat* column represents the predicted value of the outcome variable: this is essential for plotting the predicted sinusoid curve.

```{r showtoplot}

head(getKronos_trace(output))

```

4). The *getKronos_groupwise()* function arguably fetches the most useful output: this provides us with the p-value (p.val) and proportion of the variance in the data explained (r.sq) when we fit our sinusoid curve. Additionally we obtain the acrophase (acro) and amplitude of the predicted curve, which can be used in our graphics functions to visualise changes in curve with interventions (see `gg_kronos_circle`, explored further below).

```{r showindfit, echo=FALSE}

getKronos_groupwise(output)


```

### Figures

The package contains custom ggplot2 figure functions, that utilise the kronos output to rapidly produce figures that convey important information for circadian rhythms:

1.  `gg_kronos_circle()` generates a plot showing the acrophase and amplitude of the predicted curve, allowing the reader to rapidly access summary data regarding variables of interest, and to compare the summary data between groups in more complex models.At baseline, non-significant outcome measures are presented using dashed lines.
2.  `gg_kronos_sinusoid()` generates a x-y plot showing the outcome variable across the defined period. These graphs are useful for visualising the differences between specific timepoints assessed.

```{r figures}

gg_kronos_circle(output)

gg_kronos_sinusoid(output)
```

## 3. Comparing Rhythmicity for More than Two Groups

Next we will demonstrate one of the unique features of the Kronos package: the ability to compare circadian rhythms between more than two groups. This is increasingly important as the use of complex experimental designs grows in biological science. This example comprises of three independent groups and is similar in setup to a one-way ANOVA. For examples of more complex designs, see Excursion 1.

```{r figures_complex}

output2 <- kronos(formula = Variable_1 ~ Treatment + time(Timepoint), 
                  data = groupdata, period = 24, 
                  verbose = T, pairwise = T)

gg_kronos_circle(output2)

gg_kronos_sinusoid(output2)
```

There are a few changes to the output generated by the kronos function:

1.  The *getKronos_groupwise()* output now contains a line for each of our groups.

```{r showindfit_complex, echo=FALSE}

getKronos_groupwise(output2)

```

In this example you can see that groups A and B exhibit statistically significant rhythms, while the model fitted to group C is non-significant.

2.  We can now generate *pairwise models* output using `pairwise=T`. This generates pairwise comparisons between each of the groups:

```{r showpairwisemodel}

getKronos_pairwise(output2)

```

Above you can see that overall group A is significantly different between B and C, and that group B exhibits a significantly different rhythm from A and C.

3.  When including independent variables, kronos will also calculate an overall interaction with time which can be accessed using the code below:

```{r showpairwiseps}

getKronos_pairwise_p(output2)

```

This is calculated by performing a Bonferroni correction on the interactions between both the sine and cosine time components and the independent variable. The p-value reported is the lowest following correction.

## 4. O'mics Analysis

Now we will demonstrate how to adapt the package for o'mics analysis, where there are many outcome variables. We transform our data into lists so that we can automate the process.

First we need to pivot the data into long-form and then combine the metadata and data to form a single dataframe. We can achieve this using the following code:

```{r combine data}

data.long <- bigdata %>%
  pivot_longer(!Animal_ID, names_to = "Variables", values_to = "Value") %>%
  as.data.frame()

data.long <- inner_join(data.long, bigmeta, by ="Animal_ID")

```

This will generate the following dataframe:

```{r show data.long}

head(data.long)

```

We then transform the dataframe into a list so that we can automate the analysis:

```{r make list, echo=FALSE, message=FALSE, warning=FALSE, cache=FALSE}

data_names <- unique(data.long$Variables) #collect all outcome variables

data.list <- vector(mode = "list", length = length(data_names)) #Create an empty container list of the appropriate length

for(n in 1:length(data_names)){
  data.list[[n]] <- data.long %>% filter(Variables == data_names[n])
}

```

Here the for-loop creates a list where every list object is the data for one of our outcome variables. Now we can analyse each outcome variable with the kronos function, using the `lapply` function, which allows us to apply a specific function to list objects and returns a list object of the same length.

```{r kronos list, echo=FALSE, warning=FALSE, message=FALSE}


out_list = lapply(X = data.list, FUN = function(y){kronos(data = y, Value ~ Group + time(Timepoint), period = 24, pairwise = T, verbose = F)})

```

Now we have a list of kronosOut objects, which contain all our results. While this can be accessed in R, this is unwieldy for result reporting and interpretation. Therefore, we recommend using another for-loop to extract your data of interest:

```{r result list, warning=FALSE, message=FALSE}

names(out_list) <- data_names

fit_list <- vector(mode = "list", length = length(data_names)) #Create an empty container list of the appropriate length

#The for-loop below generates a list containing the individual fit results
for(m in 1:length(out_list)){
  fit_list[[m]] <- getKronos_groupwise(out_list[[m]])
  names(fit_list)[m] <- names(out_list)[m]
  
}

fit_df = do.call(rbind, fit_list) #Here we collapse the list by binding the rows together

fit_df$q.val = p.adjust(fit_df$p.val, method = "BH") #Here we perform an FDR correction on the data to account for multiple tests

write.csv(fit_df, "README_files/RhythmicityResults.csv")

```

This will generate a csv containing the individual rhythmicity calculations for the whole data set with an FDR correction to account for multiple tests.

The resulting csv can be found [here](https://github.com/thomazbastiaanssen/kronos/tree/main/README_files/RhythmicityResults.csv).

```{r show fit.df}

head(fit_df)

```

We can use a similar approach to obtain other components of the kronosOut objects. Below we include the code to obtain the pairwise comparisons as a single csv, as this is slightly more difficult.

```{r pairwise list, warning=FALSE, message=FALSE}

pairwise_list = vector(mode = "list", length = length(out_list)) #Create an empty container list of the appropriate length

#The for-loop below generates a list containing the pairwise test results
for(m in 1:length(out_list)){
  pairwise_list[[m]] <- out_list[[m]]@pairwise_models
  names(pairwise_list)[m] <- names(out_list)[m]
  
}

#Generate a single bound list
bound_list <- lapply(X = pairwise_list, FUN = function(x){do.call(rbind, x)})

#collapse the bound list for each variable into a single dataframe 
pairwise_df <- do.call(rbind, bound_list)

#separate the comparison and the effects to make results more readable
pairwise_csv = pairwise_df %>% 
  rownames_to_column("ID") %>%
  separate(col = ID, into = c("Feature","ID"), sep = "\\.", extra = "merge") %>%
  separate(col = ID, into = c("Comparison","Effect"), sep = "\\.")
  

write.csv(pairwise_csv, "README_files/PairwiseResults.csv") 


```

The resulting csv can be found [here](https://github.com/thomazbastiaanssen/kronos/tree/main/README_files/PairwiseResults.csv).

### Figures

`gg_kronos_acrogram()` is a visualisation function we have designed specifically for omics datasets. This function provides a polar histogram of your dataset's acrophases. This allows you to compare overall rhythmicity between groups. Below we can see that a large proportion of the variables peak between ZT20-23 in Group A, while the variables in Groups B and C are less synchronous.

```{r acrogram, warning=FALSE, message=FALSE}

gg_kronos_acrogram(out_list)

```

We can also use automation to obtain individual plots for our omics dataset. Here we will demonstrate how to obtain sinusoid curves for each outcome measure in the dataset.

```{r multi-plots, warning=FALSE, message=FALSE, fig.show = "hide"}

plot_list <- vector(mode = "list", length = length(out_list)) #Create an empty container list of the appropriate length

for(q in 1:length(out_list)){
  
  #save plot into relevant position in list
  plot_list[[q]] <- gg_kronos_sinusoid(out_list[[q]]) 
  
}

# library(patchwork)
# 
# wrap_plots(plot_list) + guide_area() + plot_layout(guides = 'collect')  #& labs(caption = '') for example to remove the stats. 


#to plot & save the feature graphs to a pdf:
pdf("README_files/plots_circadian.pdf")
for (i in 1:length(plot_list)) {
  print(plot_list[[i]])
}
dev.off()

```

The resulting pdf can be found [here](https://github.com/thomazbastiaanssen/kronos/tree/main/README_files/plots_circadian.pdf). The same approach can be applied for obtaining individual circle figures.

## 5. Discussion

Here we have presented standard data for the analysis of time-of-day. Some points to consider for your data is whether you can assume a 24-hour period, and whether your data is evenly distributed over the period. Please note that you will require a minimum of three data points over your period to make use of these functions, and indeed any function using sinusoid models. Currently the kronos package is not able to estimate period; a wide range of packages are capable of determining your period if this is necessary for your research question. Please note that period estimation requires even more temporal resolution: some recommend a minimum of sampling every 2 hours over a 48-hr window ([Hughes et al., 2017](https://journals.sagepub.com/doi/full/10.1177/0748730417728663)).

This tutorial is merely a template. Depending on your experimental set-up, findings and experimental questions you may need to adjust your approach. However, as complex statistical models become more frequent in the study of circadian rhythms, functions that can incorporate more complex design than two-group comparisons are essential for advancement of the field.

We have provided figure generation functions as clear communication of results is essential to producing good and useful science. Please see below for more details for figure customisation. We hope that both aspiring and veteran bioinformaticians and circadian rhythm biologists will find our guide helpful.

## Excursion 1. Customising Figures

*To be completed*

The two figure functions, `gg_kronos_circle()` and `gg_kronos_sinusoid()`, are designed to be fully compatible with ggplot2 and therefore are fully customisable. We recommend packages ggpubr or ggprism for producing publication-grade figures (all sinusoid figures produced in our publications with kronos have been formatted using ggprism). Below is an example of some of the customisation options available through ggplot2.

```{r customise figure}

gg_kronos_circle(output2) +
  scale_fill_manual(values = c("A" = "#169B62",
                               "B" = "#FFFFFF", 
                               "C" = "#FF883E")) +
  ggtitle("Figure title")

```

We encourage users to take advantage of the extensive range of online tutorials and add-on packages available for ggplot2.

## Excursion 2. Assessing Rhythmicity in More Complex Models

One of the key features of this package is the use of a formula input, which allows for analysis of complex models. Below we will demonstrate how kronos performs when assessing data with two independent variables.

```{r two-way output}

data3 <- read_csv("2waydata.csv")

two.way.data.long <- data3 %>%
  pivot_longer(cols=starts_with("Variable_"), names_to = "Variables", values_to = "Value") %>%
  as.data.frame()

two_way_data_names <- unique(two.way.data.long$Variables) #collect all outcome variables

data.list <- vector(mode = "list", length = length(two_way_data_names)) #Create an empty container list of the appropriate length

for(n in 1:length(two_way_data_names)){
  data.list[[n]] <- two.way.data.long %>% filter(Variables == two_way_data_names[n])
}

two_way_out_list = lapply(X = data.list, FUN = function(y){kronos(data = y, Value ~ Factor_A*Factor_B + time(Timepoint), period = 24, pairwise = T, verbose = F)})

names(two_way_out_list) <- two_way_data_names

gg_kronos_sinusoid(two_way_out_list$Variable_1)

gg_kronos_sinusoid(two_way_out_list$Variable_2)

gg_kronos_sinusoid(two_way_out_list$Variable_3)

gg_kronos_sinusoid(two_way_out_list$Variable_4)

```

Here we have analysed 4 outcome variables which all show different interaction effects. Here we will go into depth examining the effects observed in Variable 1 as an example of how to interpret kronos output for more complex designs.

```{r two-way var1 graphs}



gg_kronos_sinusoid(two_way_out_list$Variable_1)

gg_kronos_circle(two_way_out_list$Variable_1)


```

1). As before, we can use the `getKronos_groupwise()` function to obtain individual rhythmicity for each group. Here you can see that 3/4 groups exhibit rhythmicity and that both conventional groups share a similar acrophase (which is illustrated in the figures above as well).

```{r showgroupwise var1}

getKronos_groupwise(two_way_out_list$Variable_1)

```

2). With the `getKronos_pairwise_p()` function we can assess the interaction of each of our experimental factors with the time component of the model: here you can see that both main effects and the interaction significantly interact with the time component.

```{r showpairwisep var1}

getKronos_pairwise_p(two_way_out_list$Variable_1)

```

3). Next we can use the `getKronos_pairwise()` function to obtain the pairwise group comparisons. This allows us to determine how each group differs from one another. For example, here you can see that Conventional+Stress and Antibiotics+Control only exhibit a significant group*Timepoint_sin interaction. This is unsurprising as the groups have the same average value but exhibit a rhythm shifted by 12 hours.

```{r showpairwise var1}

getKronos_pairwise(two_way_out_list$Variable_1)

```
