---
title: "Kronos"
author: "Sarah-Jane Leigh & Thomaz F.S. Bastiaanssen"
date: "2022-10-26"
output:
  md_document:
    variant: gfm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Please note this package is under active construction.

# Kronos

If you use this software, please cite our work:

*#include citation here with DOI*

The following document is adapted from the supplementary materials for this manuscript.

## 0. Introduction

Here, we will demonstrate how to use the Kronos package to assess circadian rhythms in biological data collected over the day. For this demonstration we have adapted some data from our laboratory, currently undergoing peer review (reference to come).

We will demonstrate three common examples of experimental design currently encountered in circadian rhythms analysis:

1.  Circadian rhythm analysis of 1 variable over a 24-hour period

2.  Circadian rhythm analysis of 1 variable over a 24-hour period with two or more treatment groups

3.  Circadian rhythm analysis of omics data over a 24-hour period with two or more treatment groups

All R code used to transform, reorganise and plot the data is also shown below to hopefully provide a toolkit for aspiring and veteran bioinformaticians alike. It should be noted that the analysis performed here may not perfectly correspond to that performed in the published manuscript: some variables in the datasets provided have been manipulated to better demonstrate functions of the package.

*At the end of this document, we have included a few excursions into more advanced subjects we find useful, but that did not necessarily fit in with the mainline analysis.*

### Code chunk: Install Kronos and load our libraries

```{r load, message=FALSE}

#Install kronos
library(devtools)
#install_github("thomazbastiaanssen/kronos") 

#Load relevant packages
library(kronos)
library(tidyverse)
library(ggplot2)

#Load prepared data stored in Kronos library
###Thomaz to help wrangle this

```

\newpage

### Code chunk: Load our data tables

Data should be prepared in "long form" for use in Kronos; that is with values repeating in the "Timepoint" column, which defines when data was collected in a 24-hour cycle.

Our package includes example datasets that we will use in this tutorial that are pre-formatted. You can rearrange your data into long form using pivot_longer() or gather() in tidyr.

Since we're using prepared data, we already loaded it using 'data*(INSERT DATA NAME HERE)*' but typically we would do something like this:

```{r readdata, message=FALSE, cache=TRUE}

data("kronos_demo")
data1 <-   onevariable #One variable, one group
data2 <-   groupdata #One variable, more than two groups
bigdata <- bigdata #Omics dataset, more than two groups
bigmeta <- bigmeta #Omics sample metadata

```

You can see how an example of how the data is prepared below:

```{r showdata, echo=FALSE}

head(data2)

```

Here we have prepared the omics dataset with a separate metadata file as is common when working with omics datasets. A metadata file can be generated using select() in tidyr, or metadata and data can be combined using inner_join if they contain an identical column (both name and contents). The omics dataset used here has been central-log transformed to account for its compositional nature (see our guide <https://arxiv.org/abs/2207.12475> for easy centred-log transformation).

## 1. Statistical Approach

*THOMAZ PLEASE WRITE SOMETHING*

## 2. Analysing Rhythmicity in a Single Group

We will start with the most simple example: analysing circadian rhythmicity in a single experimental group for one outcome variable of interest. For this we use the kronos function:

```{r kronos1group, message=FALSE, warning=FALSE, cache=FALSE}

output <- kronos(formula = Variable_1 ~ time(Timepoint), 
                 data = data1, period = 24, 
                 verbose = T, pairwise = F)

```

Here we use the formula `Outcome Variable ~ time(Time Variable)`, which is the most simple model used by the Kronos function. We specify the period as 24 (this can be adjusted as appropriate for the data analysed). By selecting `verbose=T`, you will be able to see the models run by the kronos function: this becomes increasingly useful when you run more complex models. Finally we select `pairwise=F` here, as there are no groups to compare for differences in rhythms.

The kronos function creates a kronosOut object, containing several lists, which we will describe below:

1). The *input* contains the data that the model is based on, as well as the calculated cosine and sine components.

```{r showinput, echo=FALSE}

head(output@input)

```

2). The *fit* contains the key coefficients for the model.

3). The *to_plot* contains all the data required for graphing the sinusoid curve, which can either be used in our custom ggplot2 functions, or can be used in other graphing packages. *y_hat* is predicted value of the outcome variable: this is essential for plotting the predicted sinusoid curve.

```{r showtoplot, echo=FALSE}

head(output@to_plot)

```

4). The *ind_fit* is arguably the most useful output: this provides us with the p-value (p.val) and proportion of the variance in the data explained (r.sq) when we fit our sinusoid curve. Additionally we obtain the acrophase (acro) and amplitude of the predicted curve, which can be used in our graphics functions to visualise changes in curve with interventions (explored further in Sections 3 and 4).

```{r showindfit, echo=FALSE}

output@ind_fit

```

### Figures
```{r figures}

gg_kronos_circle(output)

gg_kronos_sinusoid(output)
```

## 3. Comparing Rhythmicity for More than Two Groups

Next we will demonstrate one of the unique features of the Kronos package: the ability to compare circadian rhythms between more than two groups. This is increasingly important as the use of complex experimental designs has increased in biological science. This example comprises of three independent groups and is similar in setup to a one-way ANOVA. For examples of more complex designs, see Excursion 1.

### Figures

### Figures
```{r figures_complex}

output2 <- kronos(formula = Variable_1 ~ Treatment + time(Timepoint), 
                  data = data2, period = 24, 
                  verbose = T, pairwise = F)

gg_kronos_circle(output2)

gg_kronos_sinusoid(output2)
```

## 4. Omics Analysis

BLEURGH

### Figures

## 5. Discussion

BLEURGH

## Excursion 1. Assessing Rhythmicity in More Complex Models
